---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


1. Download and unzip the dataset

```{r}
#make code visible
echo = TRUE
#set up directory for graphs
knitr::opts_chunk$set(fig.path='figure/')

fileUrl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destFile <- "repdata_data_activity.zip"
download.file(fileUrl, destFile, method = "auto", quiet = T)
unzip(destFile)
```

2. Load data to a data frame

```{r}
dataFrame <- read.csv("activity.csv", na.strings = "NA")
dataFrame$date <- as.Date(dataFrame$date)
```


## What is mean total number of steps taken per day?


1. Calculate the total number of steps taken per day

```{r}
stepsPerDay <- aggregate(list(steps = dataFrame$steps), list(date = dataFrame$date), sum)
```

2. Make a histogram of the total number of steps taken each day

```{r}
hist(stepsPerDay$steps,
     col = "red",
     main = "Total number of steps taken each day",
     xlab = "Steps per day"
     )
```

3. Calculate and report the mean and median of the total number of steps taken per day


Mean of the total number of steps taken per day:
```{r}
mean(stepsPerDay$steps, na.rm = T)
```

Median of the total number of steps taken per day
```{r}
median(stepsPerDay$steps, na.rm = T)
```


## What is the average daily activity pattern?


1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
stepsByInterval <- aggregate(steps ~ interval, dataFrame, mean, na.action = na.omit)

plot(stepsByInterval$interval, stepsByInterval$steps, 
     type="l",
     main="The time series plot of the 5-minute interval",
     xlab="The 5-minute interval",
     ylab="The average number of steps"
     )
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
as.numeric(stepsByInterval[stepsByInterval$steps == max(stepsByInterval$steps),][1])
```


## Imputing missing values


1. Calculate and report the total number of missing values in the dataset

```{r}
sum(is.na(dataFrame$steps))
```

2. Create a new dataset that is equal to the original dataset but with the missing data filled in.
   The missing data is calculated as the mean for the interval.
   
```{r}
newDataFrame <- dataFrame

for (i in 1:nrow(newDataFrame)) {
    if (is.na(newDataFrame$steps[i])) {
        newDataFrame$steps[i] <- stepsByInterval$steps[stepsByInterval$interval == newDataFrame$interval[i]]
    }
}
```

3. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
imputedStepsPerDay <- aggregate(list(steps = newDataFrame$steps), list(date = newDataFrame$date), sum)
hist(imputedStepsPerDay$steps,
     col = "red",
     main = "Total number of imputed steps taken each day",
     xlab = "Imputed steps per day"
     )
```

Mean of the imputed number of steps taken per day:
```{r}
mean(imputedStepsPerDay$steps, na.rm = T)
```

Median of the imputed number of steps taken per day
```{r}
median(imputedStepsPerDay$steps, na.rm = T)
```

We can see that the mean value is equal to the value from the first part of the assignment and the imputed mean is equal to the imputed median.
The impact of imputing missing data on the estimates of the total daily number of steps is moving the median towards mean.


## Are there differences in activity patterns between weekdays and weekends?


1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
newDataFrame$weekdays <- factor(format(newDataFrame$date, "%A"))
levels(newDataFrame$weekdays) <- list(weekday = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
                                      weekend = c("Saturday", "Sunday")
                                      )
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
avgSteps <- aggregate(list(meanOfSteps = newDataFrame$steps), 
                      list(interval = as.numeric(as.character(newDataFrame$interval)), weekdays = newDataFrame$weekdays),
                      mean
                      )
library(lattice)
with(avgSteps,
     xyplot(meanOfSteps ~ interval | weekdays,
            type = 'l',
            xlab = 'Interval',
            ylab = 'Number of steps',
            layout = c(1,2))
            )
    
```